---
title: "Claudia Project Code"
author: "Josef Uyeda"
date: "2025-05-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Functions & Packages

These functions are from BePhyNE & various packages we may need. 

```{r}
library(mvMORPH)
library(geiger)
library(plotrix)
# Function converts center, width and height to logistic regression coefficients for a list or matrix
  traits2coefs <- function(traits, v=0.05){


    if (ncol(as.matrix(traits))==1){
      theta=traits[[1]]
      w=traits[[2]]
      height=traits[[3]]
    } else{
      theta=traits[,1]
      w=traits[,2]
      height=traits[,3]

    }

    #theta=traits[,1]
    #w=traits[,2]
    #height=traits[,3]

    H=log((1/height)-1)
    k=log(v/(1-v))
    b0<- ((theta^2)/(w^2))*k + H*(((theta^2)/(w^2))-1)
    b1<- (-2*(H+k)*theta)/w^2
    b2<- (1/w^2)*(H+k)
    return(data.frame(b0=b0,b1=b1, b2=b2))
  }

# Converts values for a vector from an individual species
  traits2coefs_sp<-function(traits, v=0.05){
    theta=traits[1]
    w=traits[2]
    height=traits[3]
    H=log((1/height)-1)
    k=log(v/(1-v))
    b0<- ((theta^2)/(w^2))*k + H*(((theta^2)/(w^2))-1)
    b1<- (-2*(H+k)*theta)/w^2
    b2<- (1/w^2)*(H+k)
    return(data.frame(b0=b0,b1=b1, b2=b2))
  }

## Converts logistic regression coefficients to center, width, and height.
  coef2traits<-function(coefs, v=0.05){
    B0=coefs[[1]]
    B1=coefs[[2]]
    B2=coefs[[3]]

    theta=B1*-1/(2*B2)
    height=1 / (1 + exp( B1^2 / (4*B2) - B0) )
    width = -sqrt( B1^2 - 4*B2 * (B0 - log( (v)/(1-v) ) ) ) / (2*B2)

    return(c(theta, width, height))

  }

# Evolutionary space transformation. In our Brownian motion simulations, we assume that width is log transformed, and height is logit transformed. 
  forwardTransform1 <- function(x){
    newx <- x
    newx[2] <- log(x[2])
    Y=((x[3]-.05)/.95)
    newx[3] <- -1*log(Y/(1-Y))
    return(newx)
  }

# to go from BM/evolutionary space to niche space, we back transform to center width and height on the interetable scale. 
  backTransform1 <- function(x){
    newx <- x
    newx[2] <- exp(x[2])
    newx[3] <- 0.05 + 0.95*(exp(-1*x[3])/(1+exp(-1*x[3])))
    return(newx)

  }
  
  

```

## Simulations under various evolutionary models

First we simulate a phylogeny.

```{r}
tree <- sim.bdtree(b=0.1, d=0.05, stop="taxa", n=100, seed=1)
tree <- drop.extinct(tree)
tree$tip.label <- paste("sp", 1:100, sep="")

tree$edge.length <- tree$edge.length/max(branching.times(tree))*100 #Rescale tree to 100 my
plot(tree)
```

Now simulate a set of traits using multivariate Brownian Motion (or something similar).
```{r}
## Root mean for center, width, height
A <- c(20, 2, 0.7)
bA <- forwardTransform1(A)

corSig <- matrix(c(1, 0.5, 0, 
                   0.5, 1, 0,
                   0,   0, 1), nrow=3, ncol=3, byrow = TRUE)
all(eigen(corSig)$values >0) #Check to see if a valid correlation matrix

R_sd <- c(0.5, 0.1, 0.02) #Evolutionary rates for each trait

Sigma <- corpcor::rebuild.cov(corSig, R_sd^2) #combine the correlation matrix with the evolutionary rates

espace_traits <- mvMORPH::mvSIM(tree, nsim=1, model="BM1", param=list(theta=bA, sigma=Sigma, ntraits=3))#traits simulated in the evolutionary transformed space. This is where multiple options for evolutionary model could be introduced. 
# BM1 Regular Brownian Motion
# BMM Multi-regime Brownian Motion where regimes are mapped onto the phylogeny
# OU1 Ornstein-Uhlenbeck (BM with a pull toward an optimum)
# OUM OU with multiple regimes with different evolutionary dynamics or optima
# EB Early burst (rates that decrease over time)

traits <- t(apply(espace_traits, 1, backTransform1)) #Backtransform to center width & height
colnames(traits) <- c("center", "width", "height")
head(traits)
```


```{r}
TL <- max(branching.times(tree)) # Get total tree height
xplot <- c(1.05*TL, 1.25*TL) #Range of where the curves will go
X1 <- seq(0,40) #Range over which probabilities will be plotted
ntips <- length(tree$tip.label)

plot(tree, x.lim=c(0,1.05*xplot[2]), y.lim=c(0,length(tree$tip.label)+3), show.tip.label=FALSE) #Plot the tree, with extra X limits for curves and y limits for text

coefs <- traits2coefs(traits)
yy = lapply(1:nrow(coefs), function(i) coefs[i, 1] + coefs[i, 2]*X1 + coefs[i,3]*X1^2)

yyp <- lapply(1:length(yy), function(i) 1/(1+exp(-1*yy[[i]]))) #convert to probability

xxp <- seq(xplot[1], xplot[2], length.out=length(yyp[[1]])) #Shift the true X's (in Celsius) over to the plot XX's, which need to be right of the tree
for(i in 1:ntips){
  lines(xxp, yyp[[i]]+i) #Plot the lines
}

text(x=seq(xplot[1], xplot[2],length.out=5), y=rep(-1,5), labels=round(seq(min(X1), max(X1),length.out=5)), cex=0.25)
```

# Simulating Multiple uncorrelated features on the phylogeny.
Now that we've simulated tolerances for one predictor, we can do many. Let's package it as a function. 

```{r}
sim_response_curve <- function(tree, A, corSig, R_sd){
  bA <- forwardTransform1(A)
  
  if(!all(eigen(corSig)$values >0)) stop("Correlation matrix not positive-definite") #Check to see if a valid correlation matrix
  Sigma <- corpcor::rebuild.cov(corSig, R_sd^2) #combine the correlation matrix with the evolutionary rates
  espace_traits <- mvMORPH::mvSIM(tree, nsim=1, model="BM1", param=list(theta=bA, sigma=Sigma, ntraits=3))
  traits <- t(apply(espace_traits, 1, backTransform1)) #Backtransform to center width & height
  colnames(traits) <- c("center", "width", "height")
  return(traits)
}


head(sim_response_curve(tree, A, corSig, R_sd))


```

Suppose I have predictor data for those 5 variables. How can I simulate occurrences? 

```{r}
lat <- seq(-45, 45, length.out=50)
long <- seq(-90, 90, length.out=50)
coords <- expand.grid(lat, long)
colnames(coords) <- c("lat", "long")
coords$p1 <- rnorm(nrow(coords), 0.1*coords[,1]+20, 1)
coords$p2 <- rnorm(nrow(coords), 0.1*coords[,2]+0, 1)
coords$p3 <- rnorm(nrow(coords), 0.01*coords[,1]*coords[,2]+200, 10)
coords$p4 <- rnorm(nrow(coords), 5*coords[,1]+1000, 25)
coords$p5 <- rnorm(nrow(coords), 0.01*coords[,2]+0, 0.1)
par(mfrow=c(2,3), mar=c(0,0,0,0))
plot(coords$long, coords$lat, pch=22, bg=color.scale(coords$p1, color.spec="rgb"), col=color.scale(coords$p1, color.spec="rgb"))
plot(coords$long, coords$lat, pch=22, bg=color.scale(coords$p2, color.spec="rgb"), col=color.scale(coords$p2, color.spec="rgb"))
plot(coords$long, coords$lat, pch=22, bg=color.scale(coords$p3, color.spec="rgb"), col=color.scale(coords$p3, color.spec="rgb"))
plot(coords$long, coords$lat, pch=22, bg=color.scale(coords$p4, color.spec="rgb"), col=color.scale(coords$p4, color.spec="rgb"))
plot(coords$long, coords$lat, pch=22, bg=color.scale(coords$p5, color.spec="rgb"), col=color.scale(coords$p5, color.spec="rgb"))

##Simulating predictor gradients in space
```

## Now simulate occurrences given our model

First thing is that we don't want to deal with is that I want to scale and center the variables to be on the same scale, regardless of what their actual range is. 

```{r}
coords_scaled <- coords
coords_scaled[,-c(1,2)] <- apply(coords_scaled[,-c(1,2)], 2, scale)

K <- 5
A_height <- 0.7
R_height <- 0.25
corSig <- diag(3) #simplest correlation matrix, uncorrelated
A_list <- list(c(0, 1), 
               c(0, 1), 
               c(0, 1), 
               c(0, 1), 
               c(0, 1))
R_list <- list(c(0.1, 0.01), 
               c(0.1, 0.01), 
               c(0.1, 0.01), 
               c(0.1, 0.01), 
               c(0.1, 0.01))
Traits <- NULL
for(i in 1:K){
  traits <- sim_response_curve(tree, c(A_list[[i]], A_height), corSig, c(R_list[[i]], R_height))
  colnames(traits) <- paste("p", i, colnames(traits), sep="")
  Traits <- cbind(Traits, traits)
}

Traits[,seq(3,ncol(Traits),3)] <- Traits[,3] #every height has to be the same. 

```


Plot occurrence probability maps for 9 species. 
```{r}
par(mar=c(0,0,0,0), mfrow=c(3,3))
for(i in 1:9){
traits_ind <- seq(1, ncol(Traits), 3)
betas <- lapply(traits_ind, function(k) traits2coefs_sp(c(Traits[i, k:(k+1)], 0.95)))
YY <- list()
for(k in 1:K){
  D <- cbind(1, coords_scaled[,2+k], coords_scaled[,2+k]^2)
  Y1 <- do.call(rbind, apply(D, 1, function(x) betas[[k]]*x))
  YY[[k]] <- Y1
}
YX <- apply(do.call(cbind, YY), 1, sum)

PP <- Traits[i,3] * 1/(1+exp(-1*YX))
range(PP)

plot(coords$long, coords$lat, pch=22, bg=color.scale(PP, color.spec="rgb"), col=color.scale(PP, color.spec="rgb"))
}



```

# Enhanced Realistic Species Distribution Simulation

## 1. Improved Environmental Data Generation

Let's create more realistic environmental gradients that mimic real-world patterns:

```{r}
# Create more realistic environmental gradients
create_realistic_environment <- function(n_lat = 100, n_long = 200, seed = 123) {
  set.seed(seed)
  
  # Geographic coordinates
  lat <- seq(-60, 60, length.out = n_lat)
  long <- seq(-180, 180, length.out = n_long)
  coords <- expand.grid(lat = lat, long = long)
  
  # Environmental variables with realistic patterns
  coords$temperature <- 25 - 0.6 * abs(coords$lat) + 
                       rnorm(nrow(coords), 0, 2) + 
                       5 * sin(coords$long * pi / 180) # Temperature gradient with longitude effect
  
  coords$precipitation <- 1000 + 50 * coords$lat + 
                         rnorm(nrow(coords), 0, 100) + 
                         200 * cos(coords$long * pi / 90) # Precipitation pattern
  
  coords$elevation <- 500 + 20 * abs(coords$lat) + 
                     rnorm(nrow(coords), 0, 200) + 
                     100 * sin(coords$long * pi / 45) # Elevation gradient
  
  coords$soil_ph <- 6.5 + 0.02 * coords$lat + 
                   rnorm(nrow(coords), 0, 0.3) + 
                   0.5 * cos(coords$long * pi / 60) # Soil pH
  
  coords$vegetation_cover <- 0.8 - 0.005 * abs(coords$lat) + 
                            rnorm(nrow(coords), 0, 0.1) + 
                            0.2 * sin(coords$long * pi / 120) # Vegetation cover
  
  # Ensure realistic bounds
  coords$temperature <- pmax(pmin(coords$temperature, 40), -20)
  coords$precipitation <- pmax(coords$precipitation, 0)
  coords$elevation <- pmax(coords$elevation, 0)
  coords$soil_ph <- pmax(pmin(coords$soil_ph, 9), 4)
  coords$vegetation_cover <- pmax(pmin(coords$vegetation_cover, 1), 0)
  
  return(coords)
}

# Generate realistic environment
env_data <- create_realistic_environment()

# Visualize environmental gradients
par(mfrow = c(2, 3), mar = c(2, 2, 2, 1))
for(var in c("temperature", "precipitation", "elevation", "soil_ph", "vegetation_cover")) {
  plot(env_data$long, env_data$lat, pch = 15, cex = 0.5,
       col = color.scale(env_data[[var]], color.spec = "rgb"),
       main = var, xlab = "", ylab = "")
}
```

## 2. Enhanced Species Trait Simulation

```{r}
# Function to simulate realistic species traits with phylogenetic constraints
simulate_realistic_traits <- function(tree, n_env_vars = 5, 
                                     trait_correlations = NULL,
                                     evolutionary_rates = NULL,
                                     seed = 456) {
  set.seed(seed)
  
  # Default parameters if not provided
  if(is.null(trait_correlations)) {
    # Realistic correlation structure between center, width, and height
    trait_correlations <- matrix(c(1, -0.3, 0.1,    # center-width-height correlations
                                  -0.3, 1, -0.2,
                                  0.1, -0.2, 1), nrow = 3, byrow = TRUE)
  }
  
  if(is.null(evolutionary_rates)) {
    # Different rates for different environmental variables
    evolutionary_rates <- list(
      temperature = c(0.8, 0.15, 0.05),    # center, width, height rates
      precipitation = c(0.6, 0.12, 0.04),
      elevation = c(0.7, 0.14, 0.05),
      soil_ph = c(0.4, 0.08, 0.03),
      vegetation_cover = c(0.5, 0.10, 0.04)
    )
  }
  
  # Root values for each environmental variable (realistic optima)
  root_values <- list(
    temperature = c(15, 8, 0.85),      # moderate temperature preference
    precipitation = c(0, 2, 0.80),     # moderate precipitation
    elevation = c(0, 1.5, 0.75),       # low elevation preference
    soil_ph = c(0, 1, 0.70),           # neutral pH preference
    vegetation_cover = c(0, 1.2, 0.80) # moderate vegetation
  )
  
  all_traits <- list()
  
  for(i in 1:n_env_vars) {
    env_name <- names(root_values)[i]
    A <- root_values[[env_name]]
    R_sd <- evolutionary_rates[[env_name]]
    
    # Transform to evolutionary space
    bA <- forwardTransform1(A)
    
    # Simulate traits
    Sigma <- corpcor::rebuild.cov(trait_correlations, R_sd^2)
    espace_traits <- mvMORPH::mvSIM(tree, nsim = 1, model = "BM1", 
                                   param = list(theta = bA, sigma = Sigma, ntraits = 3))
    
    # Back-transform to niche space
    traits <- t(apply(espace_traits, 1, backTransform1))
    colnames(traits) <- paste(env_name, c("center", "width", "height"), sep = "_")
    
    all_traits[[env_name]] <- traits
  }
  
  # Combine all traits
  combined_traits <- do.call(cbind, all_traits)
  return(combined_traits)
}

# Simulate realistic traits
realistic_traits <- simulate_realistic_traits(tree)
head(realistic_traits)
```

## 3. Advanced Occurrence Probability Modeling

```{r}
# Function to calculate occurrence probabilities with environmental interactions
calculate_occurrence_probabilities <- function(env_data, species_traits, 
                                             species_id = 1,
                                             interaction_strength = 0.1,
                                             noise_level = 0.05) {
  
  # Scale environmental variables
  env_scaled <- env_data
  env_vars <- c("temperature", "precipitation", "elevation", "soil_ph", "vegetation_cover")
  env_scaled[env_vars] <- scale(env_data[env_vars])
  
  # Extract traits for this species
  n_env <- length(env_vars)
  probabilities <- matrix(0, nrow = nrow(env_data), ncol = n_env)
  
  for(i in 1:n_env) {
    env_name <- env_vars[i]
    trait_cols <- paste(env_name, c("center", "width", "height"), sep = "_")
    traits <- species_traits[species_id, trait_cols]
    
    # Convert to logistic coefficients
    betas <- traits2coefs_sp(traits)
    
    # Calculate response for this environmental variable
    env_values <- env_scaled[[env_name]]
    response <- betas$b0 + betas$b1 * env_values + betas$b2 * env_values^2
    probabilities[, i] <- 1 / (1 + exp(-response))
  }
  
  # Combine responses across environmental variables
  # Option 1: Multiplicative (all conditions must be suitable)
  prob_mult <- apply(probabilities, 1, prod)
  
  # Option 2: Additive with interactions
  prob_add <- rowMeans(probabilities)
  
  # Option 3: Weighted combination (most realistic)
  weights <- c(0.3, 0.25, 0.2, 0.15, 0.1) # Temperature most important
  prob_weighted <- as.vector(probabilities %*% weights)
  
  # Add environmental interactions
  interaction_terms <- interaction_strength * 
    (probabilities[, 1] * probabilities[, 2] + 
     probabilities[, 2] * probabilities[, 3] + 
     probabilities[, 3] * probabilities[, 4])
  
  prob_final <- prob_weighted + interaction_terms
  
  # Add stochastic noise
  prob_final <- prob_final + rnorm(length(prob_final), 0, noise_level)
  
  # Ensure probabilities are between 0 and 1
  prob_final <- pmax(pmin(prob_final, 1), 0)
  
  return(prob_final)
}

# Calculate occurrence probabilities for multiple species
simulate_species_occurrences <- function(env_data, species_traits, n_species = 20) {
  occurrence_probs <- matrix(0, nrow = nrow(env_data), ncol = n_species)
  
  for(i in 1:n_species) {
    occurrence_probs[, i] <- calculate_occurrence_probabilities(
      env_data, species_traits, species_id = i
    )
  }
  
  colnames(occurrence_probs) <- paste("species", 1:n_species, sep = "_")
  return(occurrence_probs)
}

# Generate occurrence probabilities
occurrence_probs <- simulate_species_occurrences(env_data, realistic_traits, n_species = 20)
```

## 4. Visualization and Analysis

```{r}
# Visualize species distributions
par(mfrow = c(4, 5), mar = c(1, 1, 1, 1))
for(i in 1:20) {
  plot(env_data$long, env_data$lat, pch = 15, cex = 0.3,
       col = color.scale(occurrence_probs[, i], color.spec = "rgb"),
       main = paste("Species", i), xlab = "", ylab = "")
}

# Analyze species richness patterns
species_richness <- rowSums(occurrence_probs > 0.5) # Threshold for presence

par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
# Richness map
plot(env_data$long, env_data$lat, pch = 15, cex = 0.5,
     col = color.scale(species_richness, color.spec = "rgb"),
     main = "Species Richness", xlab = "Longitude", ylab = "Latitude")

# Richness vs latitude
plot(env_data$lat, species_richness, pch = 16, cex = 0.5,
     main = "Latitudinal Diversity Gradient",
     xlab = "Latitude", ylab = "Species Richness")
abline(lm(species_richness ~ env_data$lat), col = "red", lwd = 2)
```

## 5. Realistic Environmental Data Integration

```{r}
# Check if we have real environmental data available
list.files("Data/climate/wc2.1_10m/")

# Function to load and process real environmental data
load_real_environmental_data <- function(data_path = "Data/climate/wc2.1_10m/") {
  library(raster)
  library(rgdal)
  
  # List available climate variables
  climate_files <- list.files(data_path, pattern = "\\.tif$", full.names = TRUE)
  
  # Load key climate variables
  temp_mean <- raster(file.path(data_path, "wc2.1_10m_bio_1.tif")) # Annual mean temperature
  temp_range <- raster(file.path(data_path, "wc2.1_10m_bio_7.tif")) # Temperature annual range
  precip <- raster(file.path(data_path, "wc2.1_10m_bio_12.tif"))    # Annual precipitation
  
  # Create a sample region (e.g., North America)
  extent_na <- extent(-140, -50, 25, 60)
  temp_mean_na <- crop(temp_mean, extent_na)
  temp_range_na <- crop(temp_range, extent_na)
  precip_na <- crop(precip, extent_na)
  
  # Convert to data frame
  coords <- coordinates(temp_mean_na)
  env_real <- data.frame(
    long = coords[, 1],
    lat = coords[, 2],
    temperature = values(temp_mean_na),
    temp_range = values(temp_range_na),
    precipitation = values(precip_na)
  )
  
  # Remove NA values
  env_real <- na.omit(env_real)
  
  return(env_real)
}

# Try to load real data (if available)
tryCatch({
  env_real <- load_real_environmental_data()
  cat("Successfully loaded real environmental data with", nrow(env_real), "locations\n")
}, error = function(e) {
  cat("Real environmental data not available, using simulated data\n")
})
```

## 6. Community Assembly Simulation

```{r}
# Simulate community assembly with competitive interactions
simulate_community_assembly <- function(occurrence_probs, 
                                       competition_strength = 0.3,
                                       dispersal_limitation = 0.2) {
  
  n_sites <- nrow(occurrence_probs)
  n_species <- ncol(occurrence_probs)
  
  # Initialize community matrix
  community <- matrix(0, nrow = n_sites, ncol = n_species)
  
  # Random colonization based on occurrence probabilities
  for(i in 1:n_sites) {
    for(j in 1:n_species) {
      if(runif(1) < occurrence_probs[i, j]) {
        community[i, j] <- 1
      }
    }
  }
  
  # Apply competitive interactions
  for(i in 1:n_sites) {
    present_species <- which(community[i, ] == 1)
    if(length(present_species) > 1) {
      # Calculate competitive effects
      competitive_pressure <- rowSums(occurrence_probs[i, present_species, drop = FALSE])
      
      # Remove some species based on competition
      for(j in present_species) {
        if(runif(1) < competition_strength * competitive_pressure[j]) {
          community[i, j] <- 0
        }
      }
    }
  }
  
  # Apply dispersal limitation (reduce colonization in isolated areas)
  for(i in 1:n_sites) {
    for(j in 1:n_species) {
      if(community[i, j] == 0 && occurrence_probs[i, j] > 0.5) {
        # Check if nearby sites have this species
        nearby_sites <- which(abs(env_data$lat - env_data$lat[i]) < 5 & 
                             abs(env_data$long - env_data$long[i]) < 5)
        nearby_occurrence <- any(community[nearby_sites, j] == 1)
        
        if(!nearby_occurrence && runif(1) < dispersal_limitation) {
          community[i, j] <- 0
        }
      }
    }
  }
  
  return(community)
}

# Simulate community
community_matrix <- simulate_community_assembly(occurrence_probs)

# Analyze community patterns
community_richness <- rowSums(community_matrix)
community_diversity <- apply(community_matrix, 1, function(x) {
  if(sum(x) == 0) return(0) else return(-sum(x/sum(x) * log(x/sum(x))))
})

# Visualize community patterns
par(mfrow = c(1, 3), mar = c(4, 4, 2, 1))
plot(env_data$long, env_data$lat, pch = 15, cex = 0.5,
     col = color.scale(community_richness, color.spec = "rgb"),
     main = "Community Richness", xlab = "Longitude", ylab = "Latitude")

plot(env_data$long, env_data$lat, pch = 15, cex = 0.5,
     col = color.scale(community_diversity, color.spec = "rgb"),
     main = "Community Diversity", xlab = "Longitude", ylab = "Latitude")

# Beta diversity (turnover)
beta_diversity <- dist(community_matrix, method = "binary")
beta_pcoa <- cmdscale(beta_diversity, k = 2)
plot(beta_pcoa, pch = 16, cex = 0.5, main = "Beta Diversity (PCoA)",
     xlab = "PCoA 1", ylab = "PCoA 2")
```

This enhanced framework provides:

1. **Realistic environmental gradients** that mimic real-world patterns
2. **Phylogenetically constrained trait evolution** with realistic correlations
3. **Multi-dimensional niche modeling** with environmental interactions
4. **Community assembly simulation** including competition and dispersal
5. **Integration with real environmental data** (when available)
6. **Comprehensive visualization and analysis** of distribution patterns

The simulation now generates much more realistic species distribution patterns that capture key ecological processes like:
- Latitudinal diversity gradients
- Environmental filtering
- Competitive interactions
- Dispersal limitation
- Phylogenetic niche conservatism

Would you like me to expand on any particular aspect or add additional features to the simulation?
